Astrophysical calibration of gravitational-wave detectors

### 2015-04-27 docs/cosmic_cal.tex
Some more minor changes

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW}
detectors will detect waves emitted from $O(10s)$ of compact binary
coalescences.  One such class of these cataclysmic events, the inspiral and
merger of \ac{BNS} systems will be detected out to a maximum range of $\approx
450$ Mpc. Assuming the current best estimates for the cosmological parameters,
this is equivelent to a redshift $z\approx 0.1$. As noted by~\cite{} \ac{CBC}
systems can be used as cosmiological distance markers, otherwise known as
``standard sirens'' (analogous to the \ac{EM} standard candles). Standard sirens
allow ...

@@

The technique used throughout GW research in calibrating the detectors is
carried out through a complicated system of physical manipulations of the
Fabret-Perot Michelson interferometer; where an elaborate feedback system is
used to sustain a defined measurement in arm length difference between the
moving mirrors~\cite{LIGOCal}. In general, the interferometers are set to move
at a frequency corresponding to a GW signal. An error signal is then found
through the Pound-Driver-Hall technique~\cite{Black} for mirror stabilisation
which is proportional to the difference in mirror separation of the
Fabret-Perot Michelson interferometer. The Pound-Drever-Hall technique is
essentially a control loop feedback system in which the difference output (in
GW detectors, this is the distance the mirrors move due to the GW) and the the
output set is reduced to the highest degree of certainty. This is just a brief
overview of the calibration method currently used, and is an extensive study in
itself, see~\cite{Vitale:2012} and references therein. For advanced LIGO and Virgo,
at the frequency range used in this project, the error in the calibration using
the `hardware injection method' is roughly $10\%$~\cite{Vitale:2012}. This is a
benchmark set for the error estimation using the proposed new method in this
project.~\cm{Clean this up and expand.}

\MP{The gravitational wave strain is measured through the differential arm length changes
of the interferometer via $h = \Delta L(f) / L$, where $L$ is the total arm length.
Calibration is required to relate the actual measured interferometer error signal ouptut
$e(f)$ to this differential arm length $\Delta L$. This relation
is known as the length response function, $R(f,t)$, defined such that
\begin{equation}
\Delta L(f) = R(f,t) e(f),
\end{equation}
where $R$ various slowly in time (in comparison to transient signal timescales).
Calculation of $R$ requires the measurement of various functions within a control feedback loop
(see \cite{Vitale:2012} for a brief overview of this), which are subject to
uncertainties. In this study we will assume an estimate of $R$ is available (although
in theory we could take on the role of estimating $R$ itself), but that it differs from the
truth through some unknown scale factor, $s$, so that
\begin{equation}
h_{\rm true}(f) = s h_{\rm measured}(f) = s \frac{R(f,t) e(f)}{L}.
\end{equation}
In this analysis we will simplify the situation by assuming that $s$ is a constant with
respect to frequency, but in future studies that assumption could be dropped and $s$
could take some functional form with respect to $f$.
}

@@

@@

@@

% Define the detectors and the noise epoch we will be using
For this analysis we will consider advanced generation instruments operating
at their design sensitivity.  The network we consider is the 3 detector
H1--L1--V1 configuration.

@@

\MP{Our aim is to estimate the marginal posterior probability distribution function (pdf)
for the unknown calibration scale factor. We use Bayes' theorem for which we need to define
a likelihood function and prior pdfs on the signal parameters being estimated. In this
analysis we use a nested sampling algorithm \cite{Skilling} to provide samples from the
marginal posteriors that we require.}
We use a Gaussian likelihood function given by
%
%\begin{equation}
%  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp\Bigg[ 2\Delta
%f\sum_{k>0}\frac{|\tilde{d}(f_{k}) - \tilde{h}(f_{k})|^{2}}{S(f_{k})}\Bigg]
%\end{equation}
\begin{equation}
  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} \gws_{k,i}(\vec{\theta})
  - \frac{1}{2}\left(\gws^*_{k,i}(\vec{\theta})\gws_{k,i}(\vec{\theta}) + \tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
% 
where $\Delta f$ is the frequency resolution of the data, $S_k$ is the
$k^{\rm th}$ detector's one sided noise power spectral density, and the $i$ indices
increment over frequency with $i_{f_{\rm low}}$ and $i_{f_{\rm high}}$ corresponding to
the lower and upper range in frequencies used for our
analsysis. \MP{Need to say the
range of frequencies we integrate over, which I think was 50\,Hz to 1600\,Hz from looking
at Liam's code.} The source parameters, and the unknown scale factor, have been subsumed
into a vector $\vec{\theta}$.

@@

@@

where the noise spectral density, $S^D(f)$ is not a function of distance, shown
in section~\ref{sec:pandl}. For coalescing binary neutron stars of masses
$m_{1,2} = 1.4\,M_{\odot}$ and  $D_{L}200\,Mpc$, the expected SNR is roughly 8.
The SNR was determined from $D_{L} = [10,200]\,Mpc$  using a randomly chosen
sky position which defines the antenna patterns and the GW signal. The range of
luminosity distance used in this paper represented a high level of SNR up to
the distance at which the first predicted detectable GW source; $D_{L} =
200\,Mpc$. The detector limits for the advanced era is a luminosity distance of
$\sim 450\,Mpc$. This was not initially tested as the SNR was low enough to
cause problems with computational time. It was however tested in the random sky
position simulations, see section~\ref{sec:rndsp}. The initial value for the
calibration error used in the analysis was chosen randomly.

@@

A network of detectors are used throughout GW research. Being able to show the
effect of the calibration error and recovering its uncertainty in the analysis
would help in giving a detection more confidence. In this section, the
simulations were carried out using multiple detectors; LIGO Hanford, LIGO
Livingston, and Virgo. In reality, multiple detectors are used in order to
determine the (rough) sky position of a signal. Due to their locations, if one
site detects a signal before the others, then the position of the source can be
roughly worked out. The aim was, as before, to estimate the calibration scale
factor with the least amount of uncertainty for differing levels of SNR for a
single source of GWs.

@@

\MP{We emphasise that our work has assumed that the PSD used in the likelihood function
for estimating the calibration scale factor is an accurate representation of the
PSD at the time of the observed CBC signal. In reality the PSD may be estimated from
a period close to, but not overlapping, the signal time and therefore may be slightly
different (see e.g. \cite{2013PhRvD..88h4044L}). Our result would therefore be completely correlated
with any uncertainty in the PSD estimate, but would still offer an upper limit
on the calibration uncertainty.}

@@

We would like to acknowledge the useful discussions with a whole bunch
of people.

@@

\bibliographystyle{unsrt}
\bibliography{masterbib}
\end{document}

>>>>

%\usepackage{harvard}

@@

% introduce GWs and BNS systems
It is expected that the advanced generation of interferometric \ac{GW}
detectors will detect waves emitted from $O(10s)$ of compact binary
coalescences.  One such class of these cataclysmic events, the inspiral and
merger of \ac{BNS} systems will be detected out to a maximum range of $\approx
450$ Mpc. Assuming the current best estimates for the cosmological parameters,
this is equivelent to a redshift $z\approx 0.1$. As noted by \cite{1986Natur.323..310S} \ac{CBC}
systems can be used as cosmiological distance markers, otherwise known as
``standard sirens'' (analogous to the \ac{EM} standard candles). Standard sirens
allow ...

@@

The technique used throughout GW research in calibrating the detectors is
carried out through a complicated system of physical manipulations of the
Fabret-Perot Michelson interferometer; where an elaborate feedback system is
used to sustain a defined measurement in arm length difference between the
moving mirrors~\cite{LIGOCal}. In general, the interferometers are set to move
at a frequency corresponding to a GW signal. An error signal is then found
through the Pound-Driver-Hall technique~\cite{Black} for mirror stabilisation
which is proportional to the difference in mirror separation of the
Fabret-Perot Michelson interferometer. The Pound-Drever-Hall technique is
essentially a control loop feedback system in which the difference output (in
GW detectors, this is the distance the mirrors move due to the GW) and the the
output set is reduced to the highest degree of certainty. This is just a brief
overview of the calibration method currently used, and is an extensive study in
itself, see \cite{Vitale:2012} and references therein. For advanced LIGO and Virgo,
at the frequency range used in this project, the error in the calibration using
the `hardware injection method' is roughly $10\%$ \cite{Vitale:2012}. This is a
benchmark set for the error estimation using the proposed new method in this
project.~\cm{Clean this up and expand.}

\MP{The gravitational wave strain is measured through the differential arm length, $\Delta L$, changes
of the interferometer via $h = \Delta L(f) / L$, where $L$ is the total arm length.
Calibration is required to relate the actual measured interferometer error signal ouptut
$e(f)$ to $\Delta L$. This relation is known as the length response function, $R(f,t)$,
defined such that
\begin{equation}
\Delta L(f) = R(f,t) e(f),
\end{equation}
where $R$ varies slowly in time (in comparison to transient signal timescales).
Calculation of $R$ requires the measurement of various functions within a control feedback loop
(see \cite{Vitale:2012} for a brief overview of this), which are subject to
uncertainties. In this study we will assume an estimate of $R$ is available (although
in theory we could take on the role of estimating $R$ itself), but that it differs from the
truth through some unknown scale factor, $s$, so that
\begin{equation}
h(f) = s h_{\rm m}(f) = s \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain.
In this analysis we will simplify the situation by assuming that $s$ is a constant with
respect to frequency, but in future studies that assumption could be dropped and $s$
could take some functional form, or piece-wise fit, with respect to $f$.
}

@@

\MP{We want to assess how well a calibration scale factor can be estimated
from a single observed \ac{GW} associated with a particular \ac{sGRB}. As we {\it a priori}
have no knowledge of the likely location and distance of such an event we have performed
simulations of multiple events to see the expected distribution the accuracy of the
calibration scale factor recovery.}

@@

\MP{We perform 100 simulations at each of a range of distances from 10\,Mpc out
to 200\,Mpc. For the simulations at each distance range we have allowed the
source's sky position to be randomly drawn from a uniform distribution over
the sky, the polarisation angle $\psi$ to drawn from a uniform distribution in the
range $[0, \pi/2]$, the coalescence phase to be drawn from a uniform distribution
in the range $[0, 2\pi]$, and the inclination angle drawn from a Gaussian with range XX?.
The binary system masses have been fixed as $1.4\,{\rm M}_\odot$ (based on the
very good assumption that whatever masses are chosen in these cases they will be
recovered to high accuracy) and the time of coalescence is fixed.}

@@

% Define the detectors and the noise epoch we will be using
For this analysis we will consider advanced generation instruments operating
at their design sensitivity, \MP{with noise power spectral densities taken from
\cite{2013arXiv1304.0670L}}. The network we consider is the 3 detector H1--L1--V1 configuration.

@@

\subsection{Analysis method}

\MP{Our aim is to estimate the marginal posterior probability distribution function (pdf)
for the unknown calibration scale factor. We use Bayes' theorem for which we need to define
a likelihood function and prior pdfs on the signal parameters being estimated. In this
analysis we use a nested sampling algorithm \cite{Skilling}\footnote{The code used was based
on {\tt matlabmultinest} found at \url{https://github.com/mattpitkin/matlabmultinest}.}
to provide samples from the marginal posteriors that we require.}
We use a Gaussian likelihood function given by
%
%\begin{equation}
%  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp\Bigg[ 2\Delta
%f\sum_{k>0}\frac{|\tilde{d}(f_{k}) - \tilde{h}(f_{k})|^{2}}{S(f_{k})}\Bigg]
%\end{equation}
\begin{equation}
  p(\vec{d} | \vec{\theta}, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} \gws_{k,i}(\vec{\theta})
  - \frac{1}{2}\left(\gws^*_{k,i}(\vec{\theta})\gws_{k,i}(\vec{\theta}) + \tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
% 
where $\Delta f$ is the frequency resolution of the data, $S_k$ is the
$k^{\rm th}$ detector's one sided noise power spectral density, and the $i$ indices
increment over frequency with $i_{f_{\rm low}}$ and $i_{f_{\rm high}}$ corresponding to
the lower and upper range in frequencies used for our
analsysis. \MP{Need to say the
range of frequencies we integrate over, which I think was 40\,Hz to 1600\,Hz from looking
at Liam's code.} The source parameters, and the unknown scale factor, have been subsumed
into a vector $\vec{\theta}$.

@@

\subsection{Simulations}

@@

For coalescing binary neutron stars of masses
$m_{1,2} = 1.4\,M_{\odot}$ and  $D_{L}200\,Mpc$, the expected SNR is roughly 8.
The SNR was determined from $D_{L} = [10,200]\,Mpc$  using a randomly chosen
sky position which defines the antenna patterns and the GW signal. The range of
luminosity distance used in this paper represented a high level of SNR up to
the distance at which the first predicted detectable GW source; $D_{L} =
200\,Mpc$. The detector limits for the advanced era is a luminosity distance of
$\sim 450\,Mpc$. \sout{This was not initially tested as the SNR was low enough to
cause problems with computational time. It was however tested in the random sky
position simulations, see section~\ref{sec:rndsp}. The initial value for the
calibration error used in the analysis was chosen randomly.}

@@

\sout{A network of detectors are used throughout GW research. Being able to show the
effect of the calibration error and recovering its uncertainty in the analysis
would help in giving a detection more confidence. In this section, the
simulations were carried out using multiple detectors; LIGO Hanford, LIGO
Livingston, and Virgo. In reality, multiple detectors are used in order to
determine the (rough) sky position of a signal. Due to their locations, if one
site detects a signal before the others, then the position of the source can be
roughly worked out. The aim was, as before, to estimate the calibration scale
factor with the least amount of uncertainty for differing levels of SNR for a
single source of GWs.}

@@

\MP{We emphasise that our work has assumed that the PSD used in the likelihood function
when estimating the calibration scale factor is an accurate representation of the
PSD at the time of the observed CBC signal. In reality the PSD may be estimated from
a period close to, but not overlapping, the signal time and therefore may be slightly
different \cite{2013PhRvD..88h4044L}. Our result would therefore be completely correlated
with any uncertainty in the PSD estimate, but would still offer an upper limit
on the calibration uncertainty.}

@@

We would like to acknowledge the useful discussions with a whole bunch
of people. LW was part funded for this work through a Royal Astronomical
Society Undergraduate Research Bursary.
MP is funded by the STFC under grant number ST/L000946/1.

@@

%\biaxbliographystyle{jphysicsB}
\bibliographystyle{iopart-num}
\bibliography{masterbib}
\end{document}
